# Week 4

## Continuous distributions
Defined by "Probability density functions" (PDF)
```
dnorm(x value, mean, sd)
```

mu = mean
sigma = std 

Area under probability distribution curve is 1

Can be negative

* Unlike binomial where based on trials (0..n_trials)

Ranges in continuous distribution defined by the "Cumulative density function" (CDF)
```
pnorm(value, mean sd)
```

`pnorm()` integrates the normal distr from left side (-inf) to the value given.

```{r}
pnorm(30, 40, 10)
```

What about a slice?
* ex. What fraction of dist between 25 and 30?

```{r}
pnorm(30, 40, 10) - pnorm(25, 40, 10)
```

Upper part. Greater than certain value
```{r}
1 - pnorm(30, 40, 10)
```

## Q1
```{r}
# Also rule-of-thumb (95% in two stdev), so 5% / 2 = 0.225
1 - pnorm(7, 5, 1)
```

## Rules-of-thumb for all normal distributions
* 1 stdev within mean covers 68%
* 2 stdev within mean covers 95%
* 3 stdev within mean covers 99.7%

## Q2
Less than 60 mm Hg
```{r}
pnorm(60, 77, 11.6)
```

## Q3
Greater than 90 mm Hg
```{r}
1- pnorm(90, 77, 11.6)
```

## Q4
Between 60 and 90 mm Hg
```{r}
pnorm(90, 77, 11.6) - pnorm(60, 77, 11.6)
```

## Comparison between normal and binomial distribution
Discrete steps.
Binomial cannot use floats

When `np(1-p)` is `>10`, the binomial distribution with success rate `p`
approximates to the normal distribution with `μ = np` and `σ2 = np(1 – p)`

## Comparison between other distributions
Hypergeometric

* Number of samples less than 10%
* approximates binomial with p = successes/total population and n = k

Poisson

* m > 10
* approximates to normal distribution

## Sampling and SEM
Non-trivial to get representative sample

Take many random samples of a population and calculate the mean of each sample, these means will form a normal distribution around the population 

Standard error of the mean (SEM)

* Standard deviation of this distibution
* **How well does sample represent the mean?**
    * Larger sample = more certain (precision)

```
SEM = sd / sqrt(N)
```

```{r}
data <- c(1, 2, 3)
SEM <- sd(data) / sqrt(length(data))
SEM
```

```{r}
library("plotrix")

SEM_plotrix <- std.error(data)
print(SEM == SEM_plotrix)
```

Average across n samples.

## Confidence Interval
How sure we are the population mean is within our calculated sample mean?

less confident with smaller interval

```
CI = mean +/- critical value * SEM 
```

* critical value often the stdev
* 95% confidence (1.96 stdev) is common

Need to get X% probability covered
```{r}
# Capture 99.5% of distribution
# mean 0
# stdev 1
qnorm(0.995, 0, 1)
```

Can also use t-test
```{r}
res <- t.test(data, conf.level = 0.95)
res$conf.int
```

Error bars and SE
* 1SE where 68% confident value in between 

## Summary
* stdev: How much variation exists among data points in the sample/population?
* SEM: How precise is my estimate of the population mean?
* CI: In what range of values would I expect the mean to lie?

## Central limit theorem
Data often distributed normally in nature

The distribution of sample means from any distribution has a normal distribution
