# Week 3

## Summary statistics; Discrete probability distributions

### Variance
* For each value:
    * calculate diff from mean and square it
    * Sum these squares
    * Divide by n-1
* Stays the same as n increases

### Standard deviation
* positive sqrt of variance

### Standard error
* drops as n increase

### Discrete variables
* are countable
* ex. age
* compared to continuous value like height

### Discrete probability distributions
* For every avlue of d_var there is a chance of happening (between 0.0 and 1.0)
* Sum of all events must be 1.0
* Prob of inverse of event A + prob A equals 1.0

```
# ex. 6-sided dice where event A is rolling 6
P(A) = chance of A happening
```

### Probabilty mass function (PMF)
* probability of getting single specific value from discrete distriubtion
    * Usually just asked "chance" of event
* For 4 die, PMF(12) = 0.096

### What if we want a range?
* Just add the PMFs

### Cumulative mass function (CMF)
* sum of PMF up to value
* ex: CMF(1) = PMF(0) + PMF(1)

### Mean formula
```
mean = sum(x * PMF(x))
```

If given x and PMF:
```{r}
x <- c(0, 1, 2, 3, 4)
pmf <- c(0.1, 0.3, 0.4, 0.15, 0.05)
sum(x * pmf)
```

### Binomial distribution
* trials with success rate (p) and outcome with two categories
    * n trials
    * p success rate
    * k successes

#### PMF
In R:
```{r}
# 30 trials (n)
# 0.3 success rate (p)
# 10 successes (k)
dbinom(10, 30, 0.3)
```

Examples of the binomial distribution

* Genetics – ¼ offspring have a recessive condition (p = 0.25). If you have 10 offspring, on average how many do you expect to have the condition?
* Epidemiology – A vaccine is effective in 70% of individuals (p=0.7). In a trial of
12 patients, what’s the probability exactly 10 respond?

#### Assumptions
* fixed number of trials (n)
* two possible outcomes
* constant probability of success (p)
* independent trials

#### CMF
Cumulative mass function
```{r}

# the number successes up to 9 [0, 9]
# number of trials (30)
# success rate (0.3)
1 - pbinom(9, 30, 0.3)
```

#### Mean
```
# n = 30 and p = 0.3
mean = np
variance = np(1-p)
```

### Hypergeometric distribution
Binomial without distribution
* p changes
* **Need number not ratio.**
* As population increases ...

```
# successes
# successful chances
# unsuccessful chances
# trials
dhyper(x, m, n ,k)
```

#### Example with changing probabilty
3 white balls and 3 black balls
* P(black) where you pick without replacement

```{r}
dhyper(3,3, 3, 3)
```

3 coin flips with 3 heads
* P(heads)
```{r}
0.5 * 0.5 * 0.5
```

#### Examples of hypergeometric distribution
* Genetic Screening of Populations: Sample individuals from a total population where some carry a specific allele (e.g., CF mutation). Probability that a certain number of sampled people are carriers
    * Don't sample same person.

#### CMF
```
# x = successes up to a value
phyper(x, m, n, k)
```

20 white balls or fewer
```{r}
phyper(20, 100, 100, 50)
```

40 white balls or more
```{r}
1 - phyper(39, 100, 100, 80) 
```

#### Assumptions
1. Fixed number of trials
1. Two possible outcomes per trial
1. We know the absolute number of the two outcomes we are picking from
    * Cannot use proportion as changes probability

### Poisson distribution
* Distribution that describes scattered events over period of time
* How many events happen in subset of this period?
* Based entirely on the mean

#### Examples: Birthdays
How many months are expected to have?

* no bday
* one bday
* two bday

#### PMF
```
# k = number of events
# mean (for bdays: 68 / 12 = 5.67)
dpois()
```

```{r}
# 0 bdays
dpois(0, 5.67)
# 1 bday
dpois(1, 5.67)
# 2 bday
dpois(2, 5.67)
```
#### Mean
```
mean = m
variance = m
```

#### CMF
```
# 1 month with 68 people is likely 0
CMF = ppois(number of events up to certain value, mean)
```

The 0 term useful
```
Frequency of not being exposed = f = dpois(0, 2)
Frequency of being exposed = 1 - f
```

For one cell
* Need to multiply by number of cells

#### Common assumptions
* events occur independently
* event rate is constant

### Geometric distribution
Less on subsequent events.

Chance of n non-events between event
* Highest at first event

```
p * (1 - p)^n
```

#### PMF
R function:
```
# dgeom(0, 0.1) = 0.1
# dgeom(1, 0.1) = 0.09
dgeom(failures between event, frequency of success)
```

#### Questions
What is the average spacing

#### CMF
Probability of having at least 6 people before next COVID patient
* Given 10% of population has COVID
```{r}
# pgeom(number of non-events, frequency)
1 - pgeom(5, 0.1)
```

#### Mean
Mean gap between patients
```
Mean = (1-p) / p
Variance =(1-p)/p^2
```

#### Examples of geometric distribution
* Get to success and start again
* Molecular Biology – In single-cell PCR, each amplification attempt has some
probability p of yielding a usable product. What is the typical number of failed
amplifications before the first successful one?


#### Critical assumptions
1. Repeated independent trials
1. Each trial has two possible outcomes
1. Prob of success is constant
1. Trials continue until the first success
